\MakeUppercase{Notes on Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-01 warmup.tex
Small touch-ups

\begin{abstract}
  These are some notes on our work of layered $H$-partitions for $k$-planar graphs
\end{abstract}

@@

\begin{lem}\lemlabel{k-induction} The setup:
  \begin{compactenum}
    \item Let $G_0$ be a $k$-plane graph.
    \item Let $G=(V(G_0), E(G))$ be some subgraph of $G_0$ (so $E(G)\subseteq E(G_0)$).
    \item Let $G^+$ be the plane graph obtained by adding a vertex at each crossing point in $G_0$.
    \item Let $T$ be a spanning tree of $G^+$ rooted at some vertex $r$.
    \item For each integer $i\ge 0$, let $V^+_i=\{v\in V(G^+):\dist_{G^+}(r,v)=i\}$, and let $V_i=V(G)\cap V^+_i$.
    \item Let $C$ be a cycle in $G^+$ with $r$ in the exterior of $C$ and such that
    \begin{compactenum} 
      \item No edge of $C$ is crossed by any edge of $G$; and
      \item $V(C)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $C[P_i]$ is a path; and
        \item $|P_i\cap V_j| \le 24k+4$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N^+$ be the subgraphs of $G$ and $G^+$ consisting only of those edges and vertices contained in $C$ or the interior of $C$.
  \end{compactenum}
  Then $N$ has an $H$-partition $(B_x : x\in V(H))$ such that
  \begin{compactenum}
    \item for every integer $j\ge 0$, and all $x\in V(H)$, $|B_x\cap V_j|\le 24k+4$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=B_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

\begin{proof}
  The proof is by induction on the number of vertices of $N^+$.

@@

  Add edges to $N^+$ to obtain a near-triangulation $G^\Delta$.  
  Sperner's Lemma ensures that $G^\Delta$ contains a triangular face $\tau=v_1v_2v_3$ whose three vertices are assigned different colours. For each $i\in\{1,2,3\}$, let $Q_i^+=P_{v_i}^+$ be the path in $T$ from $v_i$ to $v_i'$.  Let $Y^+$ denote the subgraph of $N^+$ consisting of vertices and edges of $Q_1^+$, $Q_2^+$, $Q_3^+$, and $\tau$.

@@

  For each $i\in\{1,2,3\}$, let $Z_i$ be the subset of $V(G)$ that is contained in the interior of $F_i$.  We claim that $N-X-V(C)$ does not contain any edge $vw$ with $v\in Z_i$ and $w\in Z_j$ for $i\neq j$.  Indeed, any such edge would contain a vertex of $V(Y^+)\setminus V(G)$ in its interior and therefore each of $v,w\in X\cup V(C)$.

  The set $X$ will become an element $B_x$ in our $H$-partition of $G$.  All that remains is to apply induction on three subgraphs.  For each $i\in\{1,\ldots,3\}$, let $N_i^+$ denote the subgraph of $G^+$ containing the vertices and edges of $G^+$ that are contained in the closure of $F_i$.  
  Let $G_i^-$ be the subgraph of $G$ obtained by removing the edges of $G$ that cross the boundary of $F_i$.  
  Let $N_i$ denote the subgraph of $G_i^-$ containing the vertices and edges of $G_i^-$ that are contained in the closure of $F_i$.  Note that $G_i^-$ is a subgraph of $G_0$ and that the number of vertices of $N_i^+$ is less than the number of vertices of $N^+$ since $N_i^+$ does not contain one of the vertices of $\tau$.  Therefore, we can apply induction to obtain the desired $H$-partition of $N_i$ using the original graphs $G_0$ and $G^+$, the subgraph $G_i^-$ in place of $G$, the cycle $F_i$ in place of $C$, and partition of $V(F_i)$ into $V(F_i)\cap X$, $V(F_i)\cap P_{i-1}$, and $V(F_i)\cap P_{i+1}$.

@@

\section{Remarks}

\begin{enumerate}
  \item The fact that $G'$ is a triangulation is not really important. We could arbitrarily triangulate $N'$ (compatible with $T$) and still find a Sperner triangle.
  \item The fact that $T$ is plane seems important since it lets us use Sperner's Lemma.  But maybe Sperner's Lemma is overkill. $Y^+$ is just a crappy separator of small layered width that happens to break $N$ into components for which no component is adjacent to more than 2 of $Q_1,\ldots,Q_3$.
  \item The fact that $C_i$ is a cycle of non-crossed edges in $G$ is not necessarily critical.  The non-crossed property isn't so important as long as no edge of $G-V(S^+)$ has endpoints in two different faces of $S^+$.

  More troublesome, again, is the use of Sperner's Lemma since it requires that $P_1,\ldots,P_k$ (and therefore also the boundary of $Y^+$) be not so wild that they interleave.  Again, maybe we can cook up something more robust than Sperner's Lemma.

  \item I think I can push through a proof for 2-plane graphs.
\end{enumerate}

>>>>

\begin{abstract}
  We prove that every $k$-planar graph has a layered $H$-partition of layered width $k(24k+4)$ in which $H$ has treewidth 3. This implies that $k$-planar graphs have a queue number upper-bounded by a polynomial in $k$ and have non-repetitive chromatic number upper-bounded by an exponential function of $k$.
\end{abstract}

@@

\begin{lem}\lemlabel{k-induction} The setup:
  \begin{compactenum}
    \item Let $G_0$ be a connected $k$-plane graph.
    \item Let $G=(V(G_0), E(G))$ be some subgraph of $G_0$ (so $E(G)\subseteq E(G_0)$).
    \item Let $G^+$ be the triangulation obtained by adding a vertex at each crossing point in $G_0$ and then adding additional edges to obtain a triangulation.
    \item Let $T$ be a BFS spanning tree of $G^+$ rooted at some vertex $r$.
    \item For each integer $i\ge 0$, let $V^+_i=\{v\in V(G^+):\dist_{G^+}(r,v)=i\}$, and let $V_i=V(G)\cap V^+_i$.
    \item Let $C$ be a cycle in $G^+$ with $r$ in the exterior of $C$ and such that
    \begin{compactenum} 
      \item No edge of $C$ is crossed by any edge of $G$; and
      \item $V(C)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $C[P_i]$ is a path; and
        \item $|P_i\cap V_j| \le 24k+4$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N^+$ be the subgraphs of $G$ and $G^+$ consisting only of those edges and vertices contained in $C$ or the interior of $C$.
  \end{compactenum}
  Then $N$ has an $H$-partition $(B_x : x\in V(H))$ such that
  \begin{compactenum}
    \item for every integer $j\ge 0$, and all $x\in V(H)$, $|B_x\cap V_j|\le 24k+4$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=B_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

\begin{proof}
  We remark that $N^+$ is a near-triangulation.
  The proof is by induction on the number of vertices of $N^+$.

@@

  Sperner's Lemma ensures that $N^+$ contains a triangular face $\tau=v_1v_2v_3$ whose three vertices are assigned different colours. For each $i\in\{1,2,3\}$, let $Q_i^+=P_{v_i}^+$ be the path in $T$ from $v_i$ to $v_i'$.  Let $Y^+$ denote the subgraph of $N^+$ consisting of vertices and edges of $Q_1^+$, $Q_2^+$, $Q_3^+$, and $\tau$.

@@

  For each $i\in\{1,2,3\}$, let $Z_i$ be the subset of $V(G)$ that is contained in the interior of $F_i$.  We claim that $N-X-V(C)$ does not contain any edge $vw\in E(G)$ with $v\in Z_i$ and $w\in Z_j$ for $i\neq j$.  Indeed, any such edge would contain a vertex of $V(Y^+)\setminus V(G)$ in its interior and therefore each of $v,w\in X\cup V(C)$.

  The set $X$ will become an element $B_x$ in our $H$-partition of $G$.  All that remains is to apply induction on three subgraphs.  For each $i\in\{1,\ldots,3\}$, let $N_i^+$ denote the subgraph of $G^+$ containing the vertices and edges of $G^+$ that are contained in the closure of $F_i$.  
  Let $G_i$ be the subgraph of $G$ obtained by removing the edges of $G$ that cross the boundary of $F_i$.  
  Let $N_i$ denote the subgraph of $G_i$ containing the vertices and edges of $G_i$ that are contained in the closure of $F_i$.  Note that $G_i$ is a subgraph of $G_0$ and that the number of vertices of $N_i^+$ is less than the number of vertices of $N^+$ since $N_i^+$ does not contain one of the vertices of $\tau$.  Therefore, we can apply induction to obtain the desired $H$-partition of $N_i$ using the original graphs $G_0$ and $G^+$, the subgraph $G_i$ in place of $G$, the cycle $F_i$ in place of $C$, and partition of $V(F_i)$ into $V(F_i)\cap X$, $V(F_i)\cap P_{i-1}$, and $V(F_i)\cap P_{i+1}$.

@@

% \section{Remarks}
% \begin{enumerate}
%   \item The fact that $G'$ is a triangulation is not really important. We could arbitrarily triangulate $N'$ (compatible with $T$) and still find a Sperner triangle.
%   \item The fact that $T$ is plane seems important since it lets us use Sperner's Lemma.  But maybe Sperner's Lemma is overkill. $Y^+$ is just a crappy separator of small layered width that happens to break $N$ into components for which no component is adjacent to more than 2 of $Q_1,\ldots,Q_3$.
%   \item The fact that $C_i$ is a cycle of non-crossed edges in $G$ is not necessarily critical.  The non-crossed property isn't so important as long as no edge of $G-V(S^+)$ has endpoints in two different faces of $S^+$.
%   More troublesome, again, is the use of Sperner's Lemma since it requires that $P_1,\ldots,P_k$ (and therefore also the boundary of $Y^+$) be not so wild that they interleave.  Again, maybe we can cook up something more robust than Sperner's Lemma.
%   \item I think I can push through a proof for 2-plane graphs.
% \end{enumerate}
